---
author: "Gadir Mamedov"
title: "Beta regression and panel data"
output: 
  html_notebook:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 5
    number_sections: true
    theme: lumen
---

## Aim:

- Create an output of 5 models: only fixed effects beta, no fixed effects + predictors, predictors + fixed effects (no membership), predictors + lagged variable, plm + fixed, plm + fixed + lag 
- check their heteroscedasticity and autocorrelation 
- plot predictions for the best model to be used 

## Helpers

Libraries
```{r}
library(forecast)
library(reshape)
library(betareg)
library(plm)
library(ggplot2)
library(knitr)
library(dplyr)
library(summarytools)
library(Hmisc)
library(car)
library(sjPlot)
library(lmtest)

```

Function to show output in the notebook
```{r}
show_output = function(model, robust = FALSE){
  model = model
  robust = robust
  if(robust == TRUE){
    knitr::knit_print(sjPlot::tab_model(model, show.se = TRUE, show.fstat = TRUE, show.aic = TRUE, show.loglik = TRUE, p.style = "stars", emph.p = TRUE, digits = 2, robust = TRUE, collapse.ci = TRUE))
  }else{
    
    knitr::knit_print(sjPlot::tab_model(model, show.se = TRUE, show.fstat = TRUE, show.aic = TRUE, show.loglik = TRUE, p.style = "stars", emph.p = TRUE, digits = 2, collapse.ci = TRUE))
  }
}
```

## Data 

> Data can be downloaded from this link and saved as a csv file. 

[LINK to dataset](https://raw.githubusercontent.com/gzmamedov/Thesis2021/main/mamedov_thesis_data.csv)

```{r}
df_base = read.csv("mamedov_thesis_data.csv")
```

## Summary of all variables

This dataset is a bit messy, and there is a lot of not relevant variables. 

```{r fig.width=9, fig.height=6}
print(dfSummary(df_base, valid.col = FALSE, graph.magnif = 0.75), 
      max.tbl.height = 300, method = "render")
```

## Dependent variables

1st variable - state reforms only 2nd variable - IPI-based progress retrieved from the reports 3nd variable - Control of Corruption index rescaled from 0 to 1 for the beta regression model.

```{r}
ggplot(df_base) +  
  geom_line(aes(year,progress_corr_only, col = "yellow")) + 
  geom_line(aes(year, progress_ipi, col = "red"))+ 
  geom_line(aes(year, coc_norm,col = "blue"))+ 
  facet_wrap(~country) + 
  scale_x_continuous(breaks = seq(2003,2020, 1))+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))+
  scale_color_identity(name = "Progress index",
                          breaks = c("yellow", "red", "blue"),
                          labels = c("Progress in state reforms", "IPI-based progress", "Control of Corruption Index"),
                          guide = "legend")+
  ylab("Progress")+
  xlab("Year")+
  ggtitle("Progress in the fight against corruption")
```


### Control of corruption index - original scale

```{r}
ggplot(df_base) + 
  geom_line(aes(year, e_wbgi_cce))+
  facet_wrap(~country) + 
  scale_x_continuous(breaks = seq(2003,2020, 1))+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))+
  ylab("CoC original scale")+
  xlab("Year")
```

Control of Corruption index is less volatile for majority of countries, and only Georgia has seen the growth over years. Thus we can conclude that not counting for time effect can lead to unnecessary multiplication of data. IPI-based progress index shows less association between years within countries.


## Notes before running models 

We will use two types of regression models: 

- Beta regression model (from betareg library)
- Panel regression (from plm library)

Whar R-squared is used for each model? 

- **Beta-regression**

> Pseudo-R2 formula = cor(log(y/(1 - y)), x %\*% beta)\^2

- **Panel regression**

[Source](https://stats.stackexchange.com/questions/79723/interpretation-of-r2-in-fixed-effects-panel-regression/79734#79734)

**overall ð‘…2**: that's the usual ð‘…2 which you would get from regressing your dependent variable ð‘Œð‘–,ð‘¡ on the explanatory variables ð‘‹ð‘–,ð‘¡.

**between ð‘…2**: if you collapse your data and remove the time component by taking the means of your variables for each panel unit individually, the ð‘…2 from regressing these time de-meaned data gives the between ð‘…2. That's the regression ð‘Œð‘–,. on ð‘‹ð‘–,. (where . replaces the time-subscript to show that time has been averaged out for each panel unit ð‘–). So this disregards all the within information in the data.

**within ð‘…2**: this comes from the prediction equation (ð‘ŒË†ð‘–,ð‘¡âˆ’ð‘ŒâŽ¯âŽ¯âŽ¯âŽ¯Ë†ð‘–,.)=(ð‘‹ð‘–,ð‘¡âˆ’ð‘‹âŽ¯âŽ¯âŽ¯âŽ¯âŽ¯ð‘–,.)ð›½Ë†), where ð‘ŒâŽ¯âŽ¯âŽ¯âŽ¯Ë†ð‘–,. and ð‘‹âŽ¯âŽ¯âŽ¯âŽ¯âŽ¯Ë†ð‘–,. are the grand means of your variables. So the within ð‘…2 gives you the goodness of fit measure for the individual mean de-trended data which disregards all the between information in the data.

For each type of PLM regression we get a corresponding R2. Beta regression, instead, shows only overall R2. It is misleading to compare within-R2 (which explains variance - variance explained by fixed effects) and overall R2 of the beta regression predictions. Thus, one solution is to compare overall R-squared. Basically, our idea is to get the best working model or understand if we get similar results using different models. 

## Only fixed effects models - Beta regression 

### IPI-based index

```{r}
model_ipi_only_fixed  = betareg(progress_ipi~country |country, link = "logit", data = df_base)
```

```{r}
show_output(model_ipi_only_fixed)
```

**DIAGNOSTICS**

Breush-Pagan test
```{r}
bptest(model_ipi_only_fixed)
```

```{r}
par(mfrow = c(2, 2))
plot(model_ipi_only_fixed)
```

Durbin-Watson test
```{r}
dwtest(model_ipi_only_fixed)
```

Autocorrelation plot
```{r}
checkresiduals(resid(model_ipi_only_fixed))
```


### Control of Corruption index

```{r}
model_coc_only_fixed  = betareg(coc_norm~country |country, link = "logit", data = df_base)
```

```{r}
show_output(model_coc_only_fixed)
```

**DIAGNOSTICS**

Breush-Pagan test
```{r}
bptest(model_coc_only_fixed)
```

```{r}
par(mfrow = c(2, 2))
plot(model_coc_only_fixed)
```

Durbin-Watson test
```{r}
dwtest(model_coc_only_fixed)
```

Autocorrelation plot
```{r}
checkresiduals(resid(model_coc_only_fixed))
```


## Only predictors - Beta regression 

### IPI-based index
```{r}
model_ipi_only_pred = betareg(progress_ipi~membership+assistance_man_per_capita*e_wbgi_gee+export_per_capita+visa_dialogue+v2smpolsoc+socialisation_ipi+v2x_cspart-e_wbgi_gee |membership+assistance_man_per_capita*e_wbgi_gee+export_per_capita+visa_dialogue+v2smpolsoc+socialisation_ipi+v2x_cspart-e_wbgi_gee, link = "logit", data = df_base)
```

```{r}
show_output(model_ipi_only_pred)
```

**DIAGNOSTICS**

Breush-Pagan test
```{r}
bptest(model_ipi_only_pred)
```

```{r}
par(mfrow = c(2, 2))
plot(model_ipi_only_pred)
```

Durbin-Watson test
```{r}
dwtest(model_ipi_only_pred)
```

Autocorrelation plot
```{r}
checkresiduals(resid(model_ipi_only_pred))
```


### Control of Corruption index

```{r}
model_coc_only_pred  = betareg(coc_norm~membership+assistance_man_per_capita+export_per_capita+visa_dialogue+v2smpolsoc+socialisation_coc+v2x_cspart |membership+assistance_man_per_capita+export_per_capita+visa_dialogue+v2smpolsoc+socialisation_coc+v2x_cspart, link = "logit", data = df_base)
```

```{r}
show_output(model_coc_only_pred)
```

**DIAGNOSTICS**

Breush-Pagan test
```{r}
bptest(model_coc_only_pred)
```

```{r}
par(mfrow = c(2, 2))
plot(model_coc_only_pred)
```

Durbin-Watson test
```{r}
dwtest(model_coc_only_pred)
```

Autocorrelation plot
```{r}
checkresiduals(resid(model_coc_only_pred))
```


## Fixed effects + predictors (no membership) - Beta regression 

### IPI-based index
```{r}
model_ipi_fe_pred = betareg(progress_ipi~country+assistance_man_per_capita*e_wbgi_gee+export_per_capita+visa_dialogue+v2smpolsoc+socialisation_ipi+v2x_cspart-e_wbgi_gee |country+assistance_man_per_capita*e_wbgi_gee+export_per_capita+visa_dialogue+v2smpolsoc+socialisation_ipi+v2x_cspart-e_wbgi_gee, link = "logit", data = df_base)
```

```{r}
show_output(model_ipi_fe_pred)
```

**DIAGNOSTICS**

Breush-Pagan test
```{r}
bptest(model_ipi_fe_pred)
```

```{r}
par(mfrow = c(2, 2))
plot(model_ipi_fe_pred)
```

Durbin-Watson test
```{r}
dwtest(model_ipi_fe_pred)
```

Autocorrelation plot
```{r}
checkresiduals(resid(model_ipi_fe_pred))
```


### Control of Corruption index

```{r}
model_coc_fe_pred  = betareg(coc_norm~country+assistance_man_per_capita+export_per_capita+visa_dialogue+v2smpolsoc+socialisation_coc+v2x_cspart |country+assistance_man_per_capita+export_per_capita+visa_dialogue+v2smpolsoc+socialisation_coc+v2x_cspart, link = "logit", data = df_base)
```

```{r}
show_output(model_coc_fe_pred)
```

**DIAGNOSTICS**

Breush-Pagan test
```{r}
bptest(model_coc_fe_pred)
```

```{r}
par(mfrow = c(2, 2))
plot(model_coc_fe_pred)
```

Durbin-Watson test
```{r}
dwtest(model_coc_fe_pred)
```

Autocorrelation plot
```{r}
checkresiduals(resid(model_coc_fe_pred))
```

## Correction of beta regressions with lagged variables

Autocorrelation becomes an issue for all models. A very strong autocorrelation observed in the 1 step lag. Thus, if the lag is included into the model, we can capture this part and provide more robust results. 


Let's create lag variables. NAs filled with the same value. 

```{r}
  lag_df = df_base %>%
  arrange(country, year) %>% 
  group_by(country) %>% 
  mutate(progress_ipi_lag = coalesce(Hmisc::Lag(progress_ipi), progress_ipi), coc_norm_lag = coalesce(  Hmisc::Lag(coc_norm), coc_norm), progress_ipi_logit_lag =  coalesce(Hmisc::Lag(progress_ipi_logit), progress_ipi_logit) , e_wbgi_cce_lag= coalesce(Hmisc::Lag(e_wbgi_cce), e_wbgi_cce) ) %>% 
  ungroup()
```

```{r}
lag_df[c("country", "year", "progress_ipi", "progress_ipi_lag", "coc_norm", "coc_norm_lag", "progress_ipi_logit", "progress_ipi_logit_lag", "e_wbgi_cce", "e_wbgi_cce_lag")]
```


## Fixed effects + predictors + lagged variable - Beta regression 

### IPI-based index
```{r}
model_ipi_fe_pred_lag= betareg(progress_ipi~progress_ipi_lag+country+assistance_man_per_capita*e_wbgi_gee+export_per_capita+visa_dialogue+v2smpolsoc+socialisation_ipi+v2x_cspart-e_wbgi_gee | progress_ipi_lag+country+assistance_man_per_capita*e_wbgi_gee+export_per_capita+visa_dialogue+v2smpolsoc+socialisation_ipi+v2x_cspart-e_wbgi_gee, link = "logit", data = lag_df)
```

```{r}
show_output(model_ipi_fe_pred_lag)
```

**DIAGNOSTICS**

Breush-Pagan test
```{r}
bptest(model_ipi_fe_pred_lag)
```

```{r}
par(mfrow = c(2, 2))
plot(model_ipi_fe_pred_lag)
```

Durbin-Watson test
```{r}
dwtest(model_ipi_fe_pred_lag)
```

Autocorrelation plot
```{r}
checkresiduals(resid(model_ipi_fe_pred_lag))
```


### Control of Corruption index

```{r}
model_coc_fe_pred_lag  = betareg(coc_norm~coc_norm_lag+country+assistance_man_per_capita+export_per_capita+visa_dialogue+v2smpolsoc+socialisation_coc+v2x_cspart |coc_norm_lag+country+assistance_man_per_capita+export_per_capita+visa_dialogue+v2smpolsoc+socialisation_coc+v2x_cspart, link = "logit", data = lag_df)
```

```{r}
show_output(model_coc_fe_pred_lag)
```

**DIAGNOSTICS**

Breush-Pagan test
```{r}
bptest(model_coc_fe_pred_lag)
```

```{r}
par(mfrow = c(2, 2))
plot(model_coc_fe_pred_lag)
```

Durbin-Watson test
```{r}
dwtest(model_coc_fe_pred_lag)
```

Autocorrelation plot
```{r}
checkresiduals(resid(model_coc_fe_pred_lag))
```

## PLM + fixed effects

### IPI-based index

```{r}
plm_model_ipi = plm(progress_ipi_logit~membership+visa_dialogue+ assistance_man_per_capita*e_wbgi_gee+v2x_cspart+v2smpolsoc+socialisation_ipi_logit+export_per_capita-e_wbgi_gee, df_base, index = c("country"), model = "within")
```

```{r}
show_output(plm_model_ipi)
```

```{r}
#between r2 squared - explanation between panels
print(r.squared(plm_model_ipi, model = "between"),dfcor = TRUE) 
#within r2 squared - explanation within panel
print(r.squared(plm_model_ipi, model = "within"), dfcor = TRUE) 
#pooled r2 squared - weighted
print(r.squared(plm_model_ipi, model = "pooling"), dfcor = TRUE) 
```

**DIAGNOSTICS**

Breush-Pagan test
```{r}
bptest(plm_model_ipi)
```

Durbin-Watson test
```{r}
pdwtest(plm_model_ipi)
```

Autocorrelation plot
```{r}
checkresiduals(resid(plm_model_ipi))
```

### Control of Corruption index

```{r}
plm_model_coc = plm(e_wbgi_cce~membership+visa_dialogue+ assistance_man_per_capita+v2x_cspart+v2smpolsoc+socialisation_e_wbgi_cce+export_per_capita, df_base, index = c("country"), model = "within")
```

```{r}
show_output(plm_model_coc)
```

```{r}
#between r2 squared - explanation between panels
print(r.squared(plm_model_coc, model = "between"), dfcor = TRUE) 
#within r2 squared - explanation within panel
print(r.squared(plm_model_coc, model = "within"),dfcor = TRUE) 
#pooled r2 squared - weighted
print(r.squared(plm_model_coc, model = "pooling"), dfcor = TRUE) 
```

**DIAGNOSTICS**

Breush-Pagan test
```{r}
bptest(plm_model_coc)
```

Durbin-Watson test
```{r}
pdwtest(plm_model_coc)
```

Autocorrelation plot
```{r}
checkresiduals(resid(plm_model_coc))
```



## PLM + fixed effects + lagged variable

### IPI-based index

```{r}
plm_model_ipi_lag = plm(progress_ipi_logit~progress_ipi_logit_lag+membership+visa_dialogue+ assistance_man_per_capita*e_wbgi_gee+v2x_cspart+v2smpolsoc+socialisation_ipi_logit+export_per_capita-e_wbgi_gee, lag_df, index = c("country"), model = "within")
```

```{r}
show_output(plm_model_ipi_lag)
```

```{r}
#between r2 squared - explanation between panels
print(r.squared(plm_model_ipi_lag, model = "between"), dfcor = TRUE) 
#within r2 squared - explanation within panel
print(r.squared(plm_model_ipi_lag, model = "within"), , dfcor = TRUE) 
#pooled r2 squared - weighted
print(r.squared(plm_model_ipi_lag, model = "pooling"),, dfcor = TRUE) 
```

**DIAGNOSTICS**

Breush-Pagan test
```{r}
bptest(plm_model_ipi_lag)
```

Durbin-Watson test
```{r}
pdwtest(plm_model_ipi_lag)
```

Autocorrelation plot
```{r}
checkresiduals(resid(plm_model_ipi_lag))
```


### Control of Corruption index

```{r}
plm_model_coc_lag = plm(e_wbgi_cce~e_wbgi_cce_lag+membership+visa_dialogue+ assistance_man_per_capita+v2x_cspart+v2smpolsoc+socialisation_e_wbgi_cce+export_per_capita, lag_df, index = c("country"), model = "within")
```

```{r}
show_output(plm_model_coc_lag)
```

```{r}
#between r2 squared - explanation between panels
print(r.squared(plm_model_coc_lag, model = "between"), dfcor = TRUE) 
#within r2 squared - explanation within panel
print(r.squared(plm_model_coc_lag, model = "within"), , dfcor = TRUE) 
#pooled r2 squared - weighted
print(r.squared(plm_model_coc_lag, model = "pooling"),, dfcor = TRUE) 
```

**DIAGNOSTICS**

Breush-Pagan test
```{r}
bptest(plm_model_coc_lag)
```

Durbin-Watson test
```{r}
pdwtest(plm_model_coc_lag)
```

Autocorrelation plot
```{r}
checkresiduals(resid(plm_model_coc_lag))
```


## Final results

2 models passed all tests: Beta regression with lagged variable and Panel regression with lagged variable

## IPI-index
```{r}
knitr::knit_print(sjPlot::tab_model(model_ipi_fe_pred, model_ipi_fe_pred_lag, plm_model_ipi, plm_model_ipi_lag, show.se = TRUE, show.fstat = TRUE, show.aic = TRUE, show.loglik = TRUE, p.style = "stars", emph.p = TRUE, digits = 2, collapse.ci = TRUE))
```

## Control of corruption
```{r}
knitr::knit_print(sjPlot::tab_model(model_coc_fe_pred, model_coc_fe_pred_lag, plm_model_coc, plm_model_coc_lag, show.se = TRUE, show.fstat = TRUE, show.aic = TRUE, show.loglik = TRUE, p.style = "stars", emph.p = TRUE, digits = 2, collapse.ci = TRUE))
```

## Membership opportunity 

### IPI-based index
```{r}
model_ipi_fe_pred_lag_membership= betareg(progress_ipi~progress_ipi_lag+membership |progress_ipi_lag+membership, link = "logit", data = lag_df)
```

### Control of Corruption 

```{r}
model_coc_fe_pred_lag_membership= betareg(coc_norm~coc_norm_lag+membership |coc_norm_lag+membership, link = "logit", data = lag_df)
```


```{r}
knitr::knit_print(sjPlot::tab_model(model_ipi_fe_pred_lag_membership, model_coc_fe_pred_lag_membership, show.se = TRUE, show.fstat = TRUE, show.aic = TRUE, show.loglik = TRUE, p.style = "stars", emph.p = TRUE, digits = 2, collapse.ci = TRUE))
```


## Presentable results 

In the end for this study I illustrate following model 
* fixed effects model
* fixed effects model + lag
* membership + lag

### IPI-based index

```{r}
knitr::knit_print(sjPlot::tab_model(model_ipi_fe_pred, model_ipi_fe_pred_lag, model_ipi_fe_pred_lag_membership, show.se = TRUE, show.fstat = TRUE, show.aic = TRUE, show.loglik = TRUE, p.style = "stars", emph.p = TRUE, digits = 2, collapse.ci = TRUE))
```

### Control of Corruption

```{r}
knitr::knit_print(sjPlot::tab_model(model_coc_fe_pred, model_coc_fe_pred_lag,model_coc_fe_pred_lag_membership, show.se = TRUE, show.fstat = TRUE, show.aic = TRUE, show.loglik = TRUE, p.style = "stars", emph.p = TRUE, digits = 2, collapse.ci = TRUE))
```


## Notes

No problems with membership models

```{r}
checkresiduals(resid(model_ipi_fe_pred_lag_membership))
```

```{r}
checkresiduals(resid(model_coc_fe_pred_lag_membership))
```

## Predictions 

Let's get predicted values first
```{r}
lag_df$pred_ipi =  predict(model_ipi_fe_pred_lag, lag_df)
lag_df$pred_coc =  predict(model_coc_fe_pred_lag, lag_df)
```

```{r}
ggplot(df_base) + 
  geom_line(aes(year, progress_ipi, col = "black")) + 
  geom_line(aes(year, pred_ipi, col = "red"))+
  facet_wrap(~country) + 
  scale_x_continuous(breaks = seq(2003,2020, 1))+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))+
  scale_color_identity(name = "Progress index",
                          breaks = c("black", "red"),
                          labels = c("IPI-based progress(original)", "Predicted values - Betareg"),
                          guide = "legend")+
  ylab("Progress")+
  xlab("Year")+
  ggtitle("Real and predicted values for IPI-based progress")
```

```{r}
ggplot(df_base) + 
  geom_line(aes(year, coc_norm, col = "black")) + 
  geom_line(aes(year, pred_coc, col = "red"))+
  facet_wrap(~country) + 
  scale_x_continuous(breaks = seq(2003,2020, 1))+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))+
  scale_color_identity(name = "Progress index",
                          breaks = c("black", "red"),
                          labels = c("Control of Corruption (original)", "Predicted values - Betareg"),
                          guide = "legend")+
  ylab("Progress")+
  xlab("Year")+
  ggtitle("Real and predicted values for Control of Corruption")
```

